OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2022-01-18 paper/response-to-reviewers.Rmd

**The manuscript is well written and describes a new computational tool without
computer science jargon. I think this manuscript and the method will be very
well received because it does fill a gap in the computation analysis of amplicon
sequences. A have only a few comments and suggestions as outlined below:**

Thank you for your enthusiasm!

**OTU quality: The main tool for comparing OTU quality in this manuscript is
MCC. This metric is central to de novo OTU clustering in the OptiClust approach,
and I think it greatly benefits amplicon-based approaches. The MCC metric is
briefly described on Lines 54-55. However, since it is so central to the
comparisons in the current study, I wonder if the reader would benefit from a
bit more detail and, I think the reader needs to be reminded about MCC when
interpreting the results. When reviewing Fig 1: the reader should know exactly
why an MCC of 0.91 is better than 0.78 and have a way to conceptualizing the
difference. Also, it may be helpful to remind the reader how quality was
determined (see L172).**

We describe MCC in more detail under the description of the OptiFit algorithm, 
which we expect readers would read just before looking at Figure 1.
We added a few more details to the MCC description in this section (L111-113 and
L117-118).
In the caption of Figure 1 we also added examples of the confusion matrix 
values for the first and last MCC scores (L439-443, L449-450).

@@

**Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar
quality as de novo clustering, ..." please add "with OptiClust" to be more
specific about what the comparison being made is.**

Agreed. (L155)

**Figure 1: W-Z are designated as the query sequences in the caption. Z does not
appear in the distance threshold table at the top of the figure and is not
assigned to an OTU at the end of Cycle 4. From Line 115 ("Alternatively, a
sequence will remained unassigned...") the reader can infer that sequence Z is
different enough from the others that it was designated a singleton (outside of
the distance threshold discussed in lines 111-113). If I'm correct, I encourage
the authors to make this point explicitly in Figure 1's caption.**

That's correct, we added a sentence to the caption as you suggested (L448-449).

@@

**Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)"
or similar to emphasize that OptiClust was being used for those benchmark
tests.**

In Figure 3, we show results from _de novo_ clustering with both the OptiClust
algorithm in mothur (red triangles) and VSEARCH (blue triangles). 
Changing the label to "OptiClust (_de novo_)" would not be accurate, as VSEARCH 
does not use the OptiClust algorithm.
Instead, we changed the red color label from "mothur" to "OptiClust (_de novo_)
or OptiFit" in Figure 3.

**Figure 3: The second sentence in the caption ("Each dataset...") is difficult
to follow due to its structure. I suggest rewriting to make it clear that de
novo clustering was performed with OptiClust, and 4 different reference based
trials were performed (self-split (referencing the OptiClust-ed reference
subset), referencing GG, SILVA, & RDP).**

We edited the caption to make the workflow more clear (L464-469).

### 2022-01-19 paper/response-to-reviewers.md

**Lines 141-142: “Thus, open reference OptiFit produced OTUs of very
similar quality as de novo clustering, …” please add “with OptiClust” to
be more specific about what the comparison being made is.**

Agreed. (L155)

**Figure 1: W-Z are designated as the query sequences in the caption. Z
does not appear in the distance threshold table at the top of the figure
and is not assigned to an OTU at the end of Cycle 4. From Line 115
(“Alternatively, a sequence will remained unassigned…”) the reader can
infer that sequence Z is different enough from the others that it was
designated a singleton (outside of the distance threshold discussed in
lines 111-113). If I’m correct, I encourage the authors to make this
point explicitly in Figure 1’s caption.**

That’s correct, we added a sentence to the caption as you suggested
(L448-449).

### 2022-02-03 paper/paper.Rmd
Move figures after references & cite in response
Fix typo
Hopefully mSphere fixes it in the published version soon too

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton.
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better MCC score [@westcott_opticlust_2017].
The MCC uses all values from a confusion matrix and ranges from negative one to
one, with a score of one occurring when all sequence pairs are true positives
and true negatives, a score of negative one occurring when all pairs are
false positives and false negatives, and a score of zero when there are equal
numbers of true and false assignments (i.e. no better than random guessing).
Sequence pairs that are similar to each other (i.e. within the distance
threshold) are counted as true positives if they are clustered into the same OTU,
and false negatives if they are not in the the same OTU.
Sequence pairs that are not similar to each other are true negatives if they are
not clustered into the same OTU, and false positives if they are not in the same
OTU.
Thus, a pair of sequences is considered correctly assigned when their OTU
assignment matches the OTU definition set by the distance threshold.
OptiClust iterations continue until the MCC stabilizes or until a maximum number
of iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.

@@

### Data Processing Steps

@@

### Data and Code Availability

@@

## Figures

>>>>

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton.
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better MCC score [@westcott_opticlust_2017].
The MCC uses all values from a confusion matrix and ranges from negative one to
one, with a score of one occurring when all sequence pairs are true positives
and true negatives, a score of negative one occurring when all pairs are
false positives and false negatives, and a score of zero when there are equal
numbers of true and false assignments (i.e. no better than random guessing).
Sequence pairs that are similar to each other (i.e. within the distance
threshold) are counted as true positives if they are clustered into the same OTU,
and false negatives if they are not in the the same OTU.
Sequence pairs that are not similar to each other are true negatives if they are
not clustered into the same OTU, and false positives if they are in the same
OTU.
Thus, a pair of sequences is considered correctly assigned when their OTU
assignment matches the OTU definition set by the distance threshold.
OptiClust iterations continue until the MCC stabilizes or until a maximum number
of iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.

@@

### Data processing steps

@@

### Data and code availability

@@

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

```{r figheader, eval=!include_figures, results='asis'}
cat('## Figure Captions\n')
```
